\documentclass[11pt]{article}
\usepackage[pdftex]{graphicx, color}
\usepackage{amsmath}
\usepackage{listings}
\newcommand\tab[1][1cm]{\hspace*{#1}}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
}

\usepackage{tikz}
\usetikzlibrary{automata,positioning}

\headheight 8pt \headsep 20pt \footskip 30pt
\textheight 9in \textwidth 6.5in
\oddsidemargin 0in \evensidemargin 0in
\topmargin -.35in

\lstset{basicstyle=\small\ttfamily,breaklines=true}

\begin{document}
\begin{center}
\LARGE CS 143 2020 - Written Assignment 2 \\
\large Due Monday, May 4, 2020 at 11:59 PM
\end{center}
This assignment covers context free grammars and parsing. You may discuss this assignment with other students and work on the problems together. However, your write-up should be your own individual work, and you should indicate in your submission who you worked with, if applicable. Assignments can be submitted electronically through Gradescope as a PDF by 11:59 PM PDT. A \LaTeX \ template for writing your solutions is available on the course website.

\begin{enumerate}
  % Problem 1
  \item  Give the context-free grammar (CFG) for each of the following languages. Any grammar is acceptable - including ambiguous grammars -  as long as it has the correct language.
  \begin{enumerate}
    \item The set of all strings over the alphabet $\{2, -, + \}$ representing valid arithmetic expressions where each integer in the expression is a single digit and the expression evaluates to some value $\geq$ 0. \\ 
    Example Strings in the Language: 
    \begin{center}
    2+2 \tab \tab 2-2+2 \tab \tab -2-2+2+2  
    \end{center}
    Strings not in the Language: 
    \begin{center}
        +2 \tab \tab -2  \tab \tab -2+22 \tab \tab 2++2-2 \tab \tab $\epsilon$ 
    \end{center}
    
    \item  The set of all strings over the alphabet $\{string|\ |\%s|arg|"|,\}$ representing valid arguments to the c printf() function. For the purposes of this problem, treat $string$ and $arg$ as tokens of your language where $string$ represents an arbitrary length sequence of characters [A-Z][a-z] and $arg$ represents any arbitrary char*. printf() replaces each $\%s$ with the contents of arg. For instance, printf("Test $\%s$ $\%s$", foo, bar) will print "Test (contents of foo) (contents of bar)". See \href{http://www.cplusplus.com/reference/cstdio/printf/}{the c printf() documentation} for further detail. Although printf() ignores unused args, your grammar should produce strings \textbf{with an equal number of \%s and arg tokens}. Note that ',' and ' ' are in the alphabet. \\ \\ 
    Example Strings in the Language (surrounded by printf() for clarity, do not inclue printf() in the grammar): 
    \begin{center}
    printf( $""$ ) \tab \tab  printf($"string\ \%s\%s", arg, arg$) \\  printf($"\tab \%s \tab stringstring\tab"\tab,arg$\tab)
    \end{center}
    Strings not in the Language (surrounded by printf() for clarity, do not inclue printf() in the grammar): 
    \begin{center}
    printf($"\%s"$)  \tab printf($"arg \ string"$)  \tab printf($string \ \%s , \ arg$)
    \end{center} 
    
    \item  The set of all strings over the alphabet $\{0, 1\}$ in the language $L:\{0^i1^j0^k \mid j \leq i + k\}$. \\ 
    Example Strings in the Language: 
    \begin{center}
    00000 \tab \tab 000111100  \tab \tab $\epsilon$ 
    \end{center}
    Strings not in the Language: 
    \begin{center}
    1   \tab \tab 000111101
    \end{center}
    
    \newpage
    \item  The set of all strings over the alphabet $\{[,],\{,\},,\}$ which are sets. We define a set to be a collection of zero or more comma-separated arrays enclosed in an open brace and a close brace. Similarly, we define an array to be a collection of zero or more comma-separated sets enclosed in an open bracket and a close bracket. \textbf{Note} that "," is in the alphabet.\\
    Example Arrays:
    \begin{center}
    $[\{\},\{\}]$\tab $[]$ \tab $[\{[],[]\}]$   
    \end{center}
    Example Sets:
    \begin{center}
    $\{[]\}$ \tab $\{\}$ \tab $\{[\{\}], []\}$  
    \end{center}
    Example Strings in the Language: 
    \begin{center}
    $\{\}$\tab $\{[],[\{[]\}]\}$\tab $\{[\{\}, \{\}, \{\}],[]\}$   
    \end{center}
    Strings not in the Language: 
    \begin{center}
    []  \tab \tab $\{\{\}\}$ \tab \tab $\{[[]]\}$
    \end{center}
  \end{enumerate}
  
  % Problem 2
  \item  \begin{enumerate}
      \item Left factor the following grammar:
        \begin{equation*}
        \begin{split}
          S &\to I \mid I-J \mid I+K  \\
          I &\to (J-K) \mid (J)  \\
          J &\to K1 \mid K2 \\
          K &\to K3 \mid \epsilon 
        \end{split}
      \end{equation*}
      \item Eliminate left recursion from the following grammar:
        \begin{equation*}
        \begin{split}
          S &\to STS \mid ST \mid T  \\
          T &\to Ta \mid Tb \mid U\\
          U &\to T \mid c \\
        \end{split}
      \end{equation*}
  \end{enumerate}
  % Problem 3
  \item Consider the following CFG, where the set of terminals is $\{a, b, \#, \%, !\}$:
  \begin{equation*}
    \begin{split}
      S &\to \%aT\mid U!\\
      T &\to aS \mid baT \mid \epsilon \\
      U &\to \#aTU \mid \epsilon \\
    \end{split}
  \end{equation*}
  \begin{enumerate}
    \item Construct the FIRST sets for each of the nonterminals.
    \item Construct the FOLLOW sets for each of the nonterminals.
    \item  Construct the LL(1) parsing table for the grammar.
    \item Show the sequence of stack, input and action configurations that occur during an LL(1) parse of the string “\#abaa\%aba!”. At the beginning of the parse, the stack should contain a single S.

  \end{enumerate}
  
  % Problem 4
  \item What advantage does left recursion have over right recursion in shift-reduce parsing?\\
\textbf{Hint:} Consider left and right recursive grammars for the language a*. What happens if your input has a million a’s?

\newpage
% Problem 5 
\item Consider the Following Grammar G over the alphabet $\Sigma$ = $\{a,b,c\}$:
  \begin{equation*}
    \begin{split}
      S' &\to S \\
      S &\to Aa \\
      S &\to Bb \\
      A &\to Ac \\
      A &\to \epsilon\\
      B &\to Bc \\
      B &\to \epsilon \\
    \end{split}
  \end{equation*}
You want to implement G using an SLR(1) parser (note that we have already added the S' $\to$ S production for you). \begin{enumerate}
    \item Construct the first state of the LR(0) machine, compute the FOLLOW sets of A and B, and point out the conflicts that prevent the grammar from being SLR(1)
    \item Show modifications to production 4 (A $\to$ Ac) and production 6 (B $\to$ Bc) that make the grammar SLR(1) while having the same language as the original grammar G. Explain the intuition behind this result.
\end{enumerate} 
\end{enumerate}
\end{document}